Turbulence and Particle Acceleration in Giant Radio Haloes: the Origin of Seed Electrons

### 2016-09-06 paper.tex
Peng's corrections

One alternative is that scattering in the high-$\beta$ ICM is mediated by plasma instabilities (firehose, mirror) rather than Coulomb scattering. This vastly increases the scattering rate and reduces the mean free path of thermal particles. In this case, the fast modes damp by TTD on relativistic rather than thermal particles. The momentum diffusion coefficient in this case is then \citep{brunetti11, miniati15}:
\begin{equation}
D_{pp}^{\rm CR} = \frac{2 p^{2} \zeta}{X_{\rm CR}} k_L \frac{\langle \varv_{\rm c}^{2} \rangle^{2}}{c_{\rm s}^{3}}
\end{equation}
where $\zeta$ is an efficiency factor for the effectiveness of plasma instabilities (e.g., due to spatial or temporal intermittency), and $X_{\rm CR}=\eps_{\rm CR}/\eps_{\rm th}$ is the relative energy density of CRs. If we set $k_L \rightarrow k_{\rm A}$, $\varv_{c}^{2} \rightarrow \varv_{\rm A}^{2}$, as before, then: 
\begin{equation}
\frac{\tau_{\rm cl}}{\tau_{\rm D}} \approx 8 \frac{\zeta}{\beta X_{\rm CR}}
\end{equation}
In principle, this implies exponential sensitivity to $X_{\rm CR}$, a similar situation as exponential sensitivity to $X_{\rm tu}$. However, there is a self-limiting asymptotic behaviour: as $X_{\rm CR}$ increases due to turbulent reacceleration, damping increases, which limits the further growth of the CR energy density \citep{brunetti11}. One natural assumption is to assume that $\eps_{\rm CR}$ saturates at a level $\eps_{\rm CR} \sim \eps_{\rm turb} ({\rm MHD}) \sim \eps_{\rm B}$, in which case $X_{\rm CR} \sim \beta^{-1}$, so that: 
\begin{equation}
\frac{\tau_{\rm cl}}{\tau_{\rm D}} \approx 1.6 \left(\frac{\zeta}{0.2}\right)\,.
\label{eqn:ratios_CR_damp} 
\end{equation}
Given the host of uncertain factors which enter into $\zeta < 1$ (the intermittent nature of turbulence, small scale magnetic topology and the efficiency with which instabilities mediated by pressure anisotropies are triggered), this is potentially consistent with $\tau_{\rm cl} \sim \tau_{\rm D}$. This promising scenario should be investigated in more detail.

@@

\sout{It is interesting to reconsider surface brightness and spectral profiles
  if indeed $\tau_{\rm D} \sim \tau_{\rm cl}$, for the reasons mentioned above
  (e.g., equation \ref{eqn:ratios_CR_damp}). We show the results of adopting
  such an ansatz in Figure~\ref{fig:param_comp_tcl_tD}, where, similar to
  Figure~\ref{fig:param_comp}, we vary $X_{\rm tu}, \alpha_{\rm tu}, \alpha_{\rm
    CR, spat}$ about fiducial values, but this time we do not adopt a fixed
  $\tau_{\rm cl}$ but set $\tau_{\rm cl}=\tau_{\rm D}$.  They are remarkably
  revealing. In this case, one is much less sensitive to the properties of the
  turbulence, and more sensitive to the details of the CR seed population. This
  gives hope to the possibility that one could effectively marginalise over the
  very uncertain properties of turbulence at larger scales in merging clusters
  (which are unlikely to be more precisely constrained observationally in the
  near future) to learn something about the underlying CR population.}  \C{It is
  interesting to reconsider surface brightness and spectral profiles if indeed
  $\tau_{\rm D} \sim \tau_{\rm cl}$ for the reasons mentioned above (e.g.,
  equation~\ref{eqn:ratios_CR_damp}). We show the results of adopting such an
  ansatz in Figure~\ref{fig:param_comp_tcl_tD}, where, similar to
  Figure~\ref{fig:param_comp}, we vary $X_{\rm tu}$, $\alpha_{\rm tu}$, and
  $\alpha_{\rm CR, spat}$ about fiducial values.\footnote{\C{When $\tau_{\rm D}
      \sim \tau_{\rm cl}$ is adopted, the radial behaviour of $\tau_{\rm cl}$ is
      also modified. Because $\tau_{\rm D}$ decreases with radius (see
      e.g. Figure~\ref{fig:tauD}), $\tau_{\rm cl}$ has to decrease by the same
      amount, which results in a steeper radio profile. To compensate for this
      effect in Figure~\ref{fig:param_comp_tcl_tD}, we change $\alpha_{\rm CR,
        spat}=1.0\rightarrow0.5$ and decrease the relative CR-to-thermal energy
      within the virial radius by a factor five to $6\times10^{-4}$.}} Instead
  of adopting a fixed $\tau_{\rm cl}$, we set $\tau_{\rm D} = 1/4 \,\tau_{\rm
    cl}$. The results are remarkably revealing. In this case, we are indeed less
  sensitive to the properties of the turbulence, and more sensitive to the
  details of the CR seed population.  This gives hope to the possibility that
  one could effectively marginalise over the very uncertain properties of
  turbulence at larger scales in merging clusters (which are unlikely to be more
  precisely constrained observationally in the near future) to learn something
  about the underlying CR population.}

For instance, in the \C{third row of panels} of
Figure~\ref{fig:param_comp_tcl_tD}, we see that once turbulence exceeds a
threshold value $X_{\rm tu} \sim 0.1$, then its overall exact energy density
does not matter -- profiles simply converge to an asymptotic form as $X_{\rm
  tu}$ is increased. This is in contrast to the exponential sensitivity,
flattening in surface brightness, and increasing spectral curvature as $X_{\rm
  tu}$ is increased that we saw previously. A threshold value of $X_{\rm tu}$ is
necessary for turbulent reacceleration to overcome cooling. Once that condition
is satisfied, the condition $\tau_{\rm cl} \sim \tau_{\rm D}$ implies a fixed
amount of amplification: stronger turbulence implies faster acceleration, but
also dissipates more quickly. This leads to the asymptotic behaviour seen.

@@

Finally, in the \C{top panels of Figure~\ref{fig:param_comp_tcl_tD}}, we see
that we {\it are} sensitive to the spatial profile of CRs; it affects both the
radio surface brightness profile (a flatter CR distribution implies flatter
surface brightness profiles) and luminosity (once again, luminosity is larger
for more centrally concentrated CRs, since more secondaries are produced). This
arises because once turbulent reacceleration overcomes cooling, the condition
$\tau_{\rm cl} \sim \tau_{\rm D}$ provides a fixed amount of amplification in
each radial shell. The radio luminosity in each shell then depends linearly on
initial conditions, i.e., the profile of CR seeds. Interestingly, the
best-fitting profile for Coma results from a flat CR distribution -- i.e., the
outcome of efficient CR streaming. Matching the overall amplitude simply
involves a rescaling of $\tau_{\rm cl}/\tau_{\rm D}$ (equation
\ref{eqn:ratios_CR_damp}) or the CR content.

Obviously, the ideas in this section require further study. However, the notion that $\tau_{\rm cl}/\tau_{\rm D}$ could self-regulate around a fixed value is exciting, because it eliminates the main source of uncertainty -- the poorly unconstrained properties of turbulence, and implies that radio haloes could potentially provide more robust constraints on the underlying CR population.

@@

If instead we adopt $\tau_{\rm cl} \sim \tau_{\rm decay}$, i.e. the natural
lifetime of turbulence is its decay lifetime, then we can show that the
acceleration time and the lifetime of turbulence are coupled, simply because the
wave-particle interaction rate (which drives particle acceleration) and the
wave-wave interaction rate (which drives dissipation) are linked. If we are
mindful that Kraichnan turbulence is only applicable below the Alfven scale
$l_{\rm A}$ (where turbulent velocities $\varv < \varv_{\rm A}$), then remarkably the
ratio $\tau_{\rm cl}/\tau_{\rm D}$ is independent of properties of the
turbulence (such as its amplitude and outer scale) and only depends on plasma
parameters. Our results suggest that TTD on thermal particles may result in
overly rapid damping of turbulence, and inefficient acceleration (equation
\ref{eqn:ratios_TTD}), but if plasma instabilities scatter the thermal particles
so that the fast modes damp via TTD on the CRs, then the required
acceleration can take place (equation \ref{eqn:ratios_CR_damp}). In agreement
with other assessments, we find that if turbulence is Burgers rather than
Kraichnan, then turbulent reacceleration is ineffective, and some other
explanation for radio haloes (e.g., magnetic reconnection, \citealt{brunetti16}) is
necessary. If we adopt the ansatz $\tau_{\rm cl} \sim \tau_{\rm D}$, then we
find that above a threshold level of turbulence (necessary to overcome cooling),
radio properties are insensitive to both the amplitude and spatial profile of
turbulence, since only a fixed, asymptotic amount of amplification takes
place. On the other hand, they {\it are} sensitive to the properties of the seed
electron profile. This raises the exciting possibility that one could
marginalise over the highly uncertain properties of cluster turbulence to learn
about the underlying seed CR population. Our results suggest that studying not
just mean trends, but the relatively small scatter in radio-halo scaling
relations, may prove extremely fruitful in shedding light on ICM plasma
processes.

>>>>

One alternative is that scattering in the high-$\beta$ ICM is mediated by plasma instabilities (firehose, mirror) rather than Coulomb scattering. This vastly increases the scattering rate and reduces the mean free path of thermal particles. In this case, the fast modes damp by TTD on relativistic rather than thermal particles. The momentum diffusion coefficient in this case is then \citep{brunetti11, miniati15}:
\begin{equation}
D_{pp}^{\rm CR} = \frac{2 p^{2} \zeta}{X_{\rm CR}} k_L \frac{\langle \varv_{\rm c}^{2} \rangle^{2}}{c_{\rm s}^{3}}
\end{equation}
where $\zeta$ is an efficiency factor for the effectiveness of plasma instabilities (e.g., due to spatial or temporal intermittency), and $X_{\rm CR}=\eps_{\rm CR}/\eps_{\rm th}$ is the relative energy density of CRs. If we set $k_L \rightarrow k_{\rm A}$, $\varv_{c}^{2} \rightarrow \varv_{\rm A}^{2}$, as before, then: 
\begin{equation}
\frac{\tau_{\rm cl}}{\tau_{\rm D}} \approx 8 \frac{\zeta}{\beta X_{\rm CR}}
\end{equation}
In principle, this implies exponential sensitivity to $X_{\rm CR}$, a similar situation as exponential sensitivity to $X_{\rm tu}$. However, there is a self-limiting asymptotic behaviour: as $X_{\rm CR}$ increases due to turbulent reacceleration, damping increases, which limits the further growth of the CR energy density \citep{brunetti11}. One natural assumption is to assume that $\eps_{\rm CR}$ saturates at a level $\eps_{\rm CR} \sim \eps_{\rm turb} ({\rm MHD}) \sim \eps_{\rm B}$, in which case $X_{\rm CR} \sim \beta^{-1}$, so that: 
\begin{equation}
\frac{\tau_{\rm cl}}{\tau_{\rm D}} \approx 2.5 \left(\frac{\zeta}{0.3}\right)\,.
\label{eqn:ratios_CR_damp} 
\end{equation}
Given the host of uncertain factors which enter into $\zeta < 1$ (the intermittent nature of turbulence, small scale magnetic topology and the efficiency with which instabilities mediated by pressure anisotropies are triggered), this is potentially consistent with $\tau_{\rm cl} \sim \tau_{\rm D}$. This promising scenario should be investigated in more detail.

@@

\C{It is
  interesting to reconsider surface brightness and spectral profiles if indeed
  $\tau_{\rm D} \sim \tau_{\rm cl}$ for the reasons mentioned above (e.g.,
  equation~\ref{eqn:ratios_CR_damp}). We show the results of adopting such an
  ansatz in Figure~\ref{fig:param_comp_tcl_tD}, where, similar to
  Figure~\ref{fig:param_comp}, we vary $X_{\rm tu}$, $\alpha_{\rm tu}$, and
  $\alpha_{\rm CR, spat}$ about fiducial values.\footnote{\C{When \C{$\tau_{\rm cl}/\tau_{\rm decay} \sim$const} is adopted, the radial behaviour of $\tau_{\rm cl}$ is
      also modified. Because $\tau_{\rm D}$ decreases with radius (see
      e.g. Figure~\ref{fig:tauD}), $\tau_{\rm cl}$ has to decrease by the same
      amount, which results in a steeper radio profile. To compensate for this
      effect in Figure~\ref{fig:param_comp_tcl_tD}, we change $\alpha_{\rm CR,
        spat}=1.0\rightarrow0.5$ and decrease the relative CR-to-thermal energy
      within the virial radius by a factor five to $6\times10^{-4}$.}} Instead
  of adopting a fixed $\tau_{\rm cl}$, we set $\tau_{\rm
    cl}/\tau_{\rm D} = 2.5 $. The results are remarkably revealing. In this case, we are indeed less
  sensitive to the properties of the turbulence, and more sensitive to the
  details of the CR seed population.  This gives hope to the possibility that
  one could effectively marginalise over the very uncertain properties of
  turbulence at larger scales in merging clusters (which are unlikely to be more
  precisely constrained observationally in the near future) to learn something
  about the underlying CR population.}

For instance, in the \C{third row of panels} of
Figure~\ref{fig:param_comp_tcl_tD}, we see that once turbulence exceeds a
threshold value $X_{\rm tu} \sim 0.2$, then its overall exact energy density
does not matter -- profiles simply converge to an asymptotic form as $X_{\rm
  tu}$ is increased. This is in contrast to the exponential sensitivity,
flattening in surface brightness, and increasing spectral curvature as $X_{\rm
  tu}$ is increased that we saw previously. A threshold value of $X_{\rm tu}$ is
necessary for turbulent reacceleration to overcome cooling. Once that condition
is satisfied, the condition $\tau_{\rm cl}/\tau_{\rm D} \sim$const implies a fixed
amount of amplification: stronger turbulence implies faster acceleration, but
also dissipates more quickly. This leads to the asymptotic behaviour seen. \C{The value of this constant depends on plasma physics details, but we have shown that the value required to match observations is plausible.}

@@

Finally, in the \C{top panels of Figure~\ref{fig:param_comp_tcl_tD}}, we see
that we {\it are} sensitive to the spatial profile of CRs; it affects both the
radio surface brightness profile (a flatter CR distribution implies flatter
surface brightness profiles) and luminosity (once again, luminosity is larger
for more centrally concentrated CRs, since more secondaries are produced). This
arises because once turbulent reacceleration overcomes cooling, the condition
$\tau_{\rm cl}/\tau_{\rm D} \sim$const provides a fixed amount of amplification in
each radial shell. The radio luminosity in each shell then depends linearly on
initial conditions, i.e., the profile of CR seeds. Interestingly, the
best-fitting profile for Coma results from a flat CR distribution -- i.e., the
outcome of efficient CR streaming. The overall amplitude is set by $\tau_{\rm cl}/\tau_{\rm D}$ (equation
\ref{eqn:ratios_CR_damp}) and the CR content.

Obviously, the ideas in this section require further study. \C{For instance, the notion that fast modes appear abruptly at $l_{\rm A}$ and that they last for a time $\tau_{\rm cl}$ are clear approximations (the latter could be improved by solving the time dependent diffusion equation for the fast mode energy spectrum, see e.g. \citet{zhou90}).} However, the notion that $\tau_{\rm cl}/\tau_{\rm D}$ could self-regulate around a fixed value is exciting, because it eliminates the main source of uncertainty -- the poorly unconstrained properties of turbulence, and implies that radio haloes could potentially provide more robust constraints on the underlying CR population.

@@

If instead we adopt \C{$\tau_{\rm cl}/\tau_{\rm decay} \sim$const}, i.e. the natural
lifetime of turbulence is its decay lifetime, then we can show that the
acceleration time and the lifetime of turbulence are coupled, simply because the
wave-particle interaction rate (which drives particle acceleration) and the
wave-wave interaction rate (which drives dissipation) are linked. If we are
mindful that Kraichnan turbulence is only applicable below the Alfven scale
$l_{\rm A}$ (where turbulent velocities $\varv < \varv_{\rm A}$), then remarkably the
ratio $\tau_{\rm cl}/\tau_{\rm D}$ is independent of properties of the
turbulence (such as its amplitude and outer scale) and only depends on plasma
parameters. Our results suggest that TTD on thermal particles may result in
overly rapid damping of turbulence, and inefficient acceleration (equation
\ref{eqn:ratios_TTD}), but if plasma instabilities scatter the thermal particles
so that the fast modes damp via TTD on the CRs, then the required
acceleration can take place (equation \ref{eqn:ratios_CR_damp}). In agreement
with other assessments, we find that if turbulence is Burgers rather than
Kraichnan, then turbulent reacceleration is ineffective, and some other
explanation for radio haloes (e.g., magnetic reconnection, \citealt{brunetti16}) is
necessary. If we adopt the ansatz \C{$\tau_{\rm cl}/\tau_{\rm decay} \sim$const}, then we
find that above a threshold level of turbulence (necessary to overcome cooling),
radio properties are insensitive to both the amplitude and spatial profile of
turbulence, since only a fixed, asymptotic amount of amplification takes
place. On the other hand, they {\it are} sensitive to the properties of the seed
electron profile. This raises the exciting possibility that one could
marginalise over the highly uncertain properties of cluster turbulence to learn
about the underlying seed CR population. Our results suggest that studying not
just mean trends, but the relatively small scatter in radio-halo scaling
relations, may prove extremely fruitful in shedding light on ICM plasma
processes.
